Supplementary material

### 2022-03-08 analysis/paper/paper.Rmd
Slight adustment to figures and more writing

The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century. Shoreline dating has been fundamental to Stone Age archaeology in Norway ever since [e.g. @berg-hansen2009; @bjerck2008; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

@@

One of the more comprehensive evaluations of the relationship between radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.

The method of shoreline dating has variably been said to contribute a method for relative dating for sites located at different elevations, for offering a lower limit for the site could have been in use, or been used to date a larger number of sites assign uniform uncertainty ranges of 200 or 500 years.

@@

The employed shoreline displacement curves are all based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point (SLIP). Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient. The SLIPs from the isolaton basin method indicate the isolation from the highest astronomical tide. For the displacement curve, this is then adjusted to mean sea-level based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

The archaeological data compiled for the analysis consists of all excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number a total `r nrow(sites)` sites. Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates. Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE) with 95 % probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape. Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water. It could therefore seem that the Løvås sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. While these values remain the same, the mean and standard deviation is constrained slightly when only considering the Mesolithic dates. While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$topodist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.

An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)). While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data. However, this can at least in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:

Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation. The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered synchronous. The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%). The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2. For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material). If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.

@@

```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using dates corresponding to the site inventory. Negative values have been set to zero. The first row only includes dates older than 2500 BCE and the second row only dates older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory. A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black. These are plotted against the oldest radiocarbon-informed phase for each site in red. B) Using the method described by Parnell et al. (2008), the synchroneity of the dates is evaluated. When the range of the 95\\% HDR for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

```{r redate, echo = FALSE, fig.cap = "Shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. The range of the 95\\% HDR in black are comapred to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/redate.png"))
```

@@

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g. @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020]. This is also related to the fact that while mean sea-level is used for dating the sites, a consideration of the tidal range could possibly have implications for the site location relative to the shoreline, depending on the topography. If such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.

@@

This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g. @breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.

>>>>

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century. Shoreline dating has been fundamental to Stone Age archaeology in Norway ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The method of shoreline dating has ranged from being met with scepticism as related to the degree sites would have been consistently shore-bound, to being deemed a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer little more than a earliest possible date for when a site could have been in use. The most common application in Norway has arguably been to use the method to provide a very approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below]. Recently the method has also been used idenpendently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 200 or 500 year bins [e.g. @breivik2014; @solheim2018; @fossum2020; @nielsen2021b]. In his discussion of the method, @nordqvist1999 -- in part drawing on @johansen1963 -- argues that there can be little doubt about the general applicability of the method, but what is less clear is the level of reliability and chronological resolution that it can offer.

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; although see @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic of the Oslo fjord area is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

@@

One of the more comprehensive evaluations of the relationship between radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g. @persson2008; @schulke2020; see also ]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.

@@

The employed shoreline displacement curves are all based on the so-called isolation basin method [see e.g. @kjemperud1986; @romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point (SLIP). Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)). The resulting SLIPs indicate the isolation of the basin from the highest astronomical tide. This is adjusted to mean sea-level for the compilation of the displacement curve based on the present day tidal range, which is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well-defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

The archaeological data compiled for the analysis consists of excavated Stone Age sites with available spatial data from the coastal region between Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). These number `r nrow(sites)` sites. Of these, `r length(unique(rcarbon$site_name))` sites are associated with a total of `r nrow(rcarbon)` radiocarbon dates. Of these, in turn, `r nrow(sites_sa)` sites are related to the `r nrow(rcarb_sa)` radiocarbon dates that fall within the Stone Age (9500--1700 BCE), with 95 % probability. These sites and ^14^C-dates form the basis for the analysis. Spatial data in the form of site limits and features, as defined by the excavating archaeologists, were retrieved from local databases at the Museum of Cultural History---the institution responsible for archaeological excavations in the region. In the compiled dataset, each radiocarbon date has been associated with the site features or excavation unit from where they originate, or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what available spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as that defined during initial survey, area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hull polygons generated around the site features.

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape. While there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water. It could therefore seem that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geological reconstruction of shoreline displacement in the area.

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between sites and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was simulated to be situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. While these values remain the same, the mean and standard deviation is constrained slightly when only considering the Mesolithic dates in the seceond row of Figure \@ref(fig:results2). Furthermore, while the median for horizontal and topographic distances is only `r round(median(distances2$topodist))`m across all plots in Figure \@ref(fig:results2), it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the surrounding topography.

An exponential function has been fit to the distributions for vertical distance using maximum likelihood estimation (Figure \@ref(fig:results2)). While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data. However, this can at least in part be related to methodological factors where the accumulation of distance-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:

Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits. The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing synchroneity is given in Figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement. The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%). The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2. For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material). If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated. Another implication of the failed dating of these two sites could be that sites from the absolute earliest part of the Mesolithic should be dated using a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate, to accommodate for the apparent deviation from the shoreline for sites older than c. 6500 BCE (cf. Figure \@ref(fig:results)).

@@

```{r results2, echo = FALSE, fig.cap = "Histograms showing distance from shoreline using dates corresponding to the site inventory. Negative values have been set to zero. The first row only includes simulated results older than 2500 BCE and the second row only results older than 4000 BCE.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/results2.png"))
```

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here. A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black. These are plotted against the oldest radiocarbon-informed phase for each site in red. B) Using the method described by Parnell et al. (2008), the synchroneity of the dates is evaluated. When the range of the 95\\% HDR for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

```{r redate, echo = FALSE, fig.cap = "Re-dating of previously shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. The range of the 95\\% HDR in black are compared to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/redate.png"))
```

@@

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g. @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020]. This is also related to the fact that while mean sea-level is used for dating the sites, a consideration of the tidal range could possibly have implications for the site location relative to the shoreline, depending on the topography. If such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.

@@

This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account , which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.

### 2022-03-08 analysis/supplementary-materials/supplementary.Rmd
Slight adustment to figures and more writing

This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information. The sites are presented in alphabetical order.  excavated sites and radiocarbon dates, and spatial data for site limits and site features are available in the online repoistory for the paper at https://osf.io/7F9SU

Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.

>>>>

This document shows the simulation results for each individual site analysed in the study with some additional notes pertaining to edits to the DTM, treatment of ^14^C-dates and other relevant information. The sites are presented in alphabetical order. All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU

Stone Age sites for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.
